---
title: "Coral-Tissue-Analysis"
author: "Meron G. Tekle"
date: "2024-06-04"
output: html_document
---

# Analysis codes for Coral Tissue samples

## set the working directory for the analysis

```{r}
setwd("~/OneDrive/Desktop/minimalR/tissue_plots/")
```

## Load required packages for the analysis

```{r}
library(tidyverse)
library(qiime2R)
library(microbiome)
library(phyloseq)
library(stringr)
library(vegan)
```

## Import .qza files from qiime2 into R using the qiime2R package and create a phyloseq object

```{r}
physeq <- qza_to_phyloseq(
  features = "table.qza",
  tree = "rooted_tree.qza",
  "blast-taxonomy-results-99.qza",
  metadata = "sample-metadata.txt")

```

### use the microbiome package to create tible files for feature table and taxonomy table which will be merged later to make analysis easier down the road

```{r}
otu.tib <- otu_tibble(physeq, column.id = "FeatureID")  #retrieves our feature table
tax.tib <- tax_tibble(physeq, column.id = "FeatureID") #retrieves our taxonomic assignment table
sample_tib <- sample_tibble(physeq, column.id = "SampleID") #retrieves or sample metadata
```

### Then we will merge all these tables using shared columns such as "SampleID" and "featureID"

```{r}
dat <- otu.tib %>% 
  pivot_longer(-FeatureID, names_to = "SampleID", values_to = "count") #creates a file with the columns FeatureID(ASV id given by qiime), SampleID (Sample names) and Count(abundance of a feature per sample). Now it is easier to join with other tables
```

### Use left_join to add taxa table, metadata

```{r}
dat <- dat %>% 
  left_join(tax.tib, by="FeatureID")

dat <- dat %>% 
  left_join(sample_tib, by = "SampleID")

dat
```

### Now to we add shannon index and PCoA results to the data table, but first we import both shannon and PCoA results

```{r}
shannon <- read_qza("shannon_vector.qza") #import from qiime using qiime2R

shannon <- shannon$data %>% rownames_to_column("SampleID") #make it easier to merge to "dat" using "SampleID"

dat <- dat %>% 
  left_join(shannon, by = "SampleID") 

dat
```

### We also add a column for observed features

```{r}
obs_feat <- read_qza("observed_features_vector.qza") #import from qiime using qiime2R
```

```{r}
obs_feat <- obs_feat$data %>% rownames_to_column("SampleID") #make it easier to merge to "dat" using "SampleID"

dat <- dat %>% 
  left_join(obs_feat, by = "SampleID") 

```

### Now we do the same with PCoA results

```{r}
uwunifrac<-read_qza("unweighted_unifrac_pcoa_results.qza") #import unweighted unifrac results from qiime2

unifrac_modify<- uwunifrac$data$Vectors %>% 
  select(SampleID, PC1, PC2)  #we select only two dimensions because we cannot effectively show more dimension on paper

dat <- dat %>% 
  left_join(unifrac_modify, by = "SampleID")

dat
```

### Finally we will divide the depth value into four catagories and order them accordinglly

```{r}
#create a depth category
dat <- dat %>% 
  mutate(depth2= case_when(Depth <= 3 ~ "Shallow (1-3m)", 
                           Depth >= 4 & Depth <= 10 ~ "Deeper Shallow (4-10m)", 
                           Depth >= 11 & Depth <= 29 ~ "Intermediate (11-29m)", 
                           Depth >= 30 & Depth <= 50 ~ "Upper Mesophotic (30-50m)",
                           Depth >= 51 & Depth <= 70 ~ "Middle Mesophotic (51-70m)",
                           Depth >70 ~ "Lower Mesophotic (>70)"))

#order the catagories based on depth
dat$depth2 <- factor(dat$depth2, 
                              levels = c("Shallow (1-3m)", 
                                         "Deeper Shallow (4-10m)",
                                         "Intermediate (11-29m)", "Upper Mesophotic (30-50m)", 
                                         "Middle Mesophotic (51-70m)", 
                                         "Lower Mesophotic (>70)")) 
```

### Now we begin analysis, but before we create plots, lets create a custom theme. This will make our code cleaner

```{r}
my_theme_alpha <-  theme(panel.border = element_rect(colour = "black", fill = NA),
                     panel.background = element_blank(),
                     panel.grid = element_blank(),
                     axis.text  = element_text(size = 10), 
                     axis.text.x = element_text(angle = 0, face = "italic"),
                     axis.title = element_text(size = 20),
                     strip.background = element_blank(),
                     strip.text = element_text(size = 10),
                     axis.line.x = element_line(size = 0.5),
                     axis.line.y = element_line(size = 0.5),
                     legend.position = "right",
                     legend.text =element_text(size =6))

```

# Alpha diversity graphs

## Create a Shannon diversity Box Plot

### But before lets create a dataframe with each sample coressponding to a unique sahnnon diversity value

```{r}
dat_shannon <- dat %>% 
  group_by(SampleID, Type, Depth, Genus.y, depth2) %>% 
  summarise(Richness = unique(shannon_entropy))

dat_shannon
```

## Before ploting lets remove coral genera which have only one observation 
```{r}
target_genus <- c("Acropora", "Alveopora", "Galaxea", "Leptoseris", "Montipora",
                  "Pocillopora")
dat_shannon <- dat %>%
  mutate(alpha_genus= Genus.y == target_genus)


dat_shannon <- droplevels(dat_shannon[dat_shannon$Genus.y == "Acropora"|dat_shannon$Genus.y == "Alveopora"|
                                   dat_shannon$Genus.y == "Galaxea"|dat_shannon$Genus.y == "Leptoseris"|
                                   dat_shannon$Genus.y == "Montipora"|dat_shannon$Genus.y == "Pocillopora",])

dat_shannon <- dat_shannon %>% 
  group_by(SampleID, Type, Depth, Genus.y, depth2) %>% 
  summarise(Richness = unique(shannon_entropy))

dat_shannon
```

### Now lets plot it


```{r}

fig7a <-dat_shannon %>% 
  drop_na() %>% 
  filter(Type != "Planula",) %>%#removes values from samples not included in analysis because of sampling depth
  ggplot(aes(Genus.y, Richness)) +
  geom_boxplot(aes(fill = Genus.y),position = "dodge", show.legend = F)+
  geom_jitter()+
  xlab(NULL)+
  ylab("Shannon Diversity")+
  scale_fill_manual(values = c("#AA4488","#777711","#117744",
                               "#114477","#DDAA77","#771122"))+
  my_theme_alpha

fig7a

#ggsave("figures/shannon-box-plot-tissue-by-genus.png", height = 4, width = 5, device= "png")
```
## Plot shannon box for Acropora and Galaxea
```{r}
fig7c <-dat_shannon %>% 
  drop_na() %>% 
  filter(Type != "Planula", Genus.y == "Acropora"| Genus.y == "Galaxea",) %>%   #removes values from samples not included in analysis because of sampling depth
  ggplot(aes(Genus.y, Richness)) +
  geom_boxplot(aes(fill = Genus.y),position = "dodge", show.legend = F)+
  geom_jitter()+
  facet_wrap(~depth2, scales = "free_x")+
  xlab(NULL)+
  ylab("Shannon Diversity")+
  scale_fill_manual(values = c("#AA4488","#777711","#117744",
                               "#114477","#DDAA77","#771122"))+
  my_theme_alpha

fig7c

#ggsave("figures/shannon-box-plot-Acro-Gala.png", height = 4, width = 5, device= "png")
```


## Create an observed features box plot

### create a custom theme for the plot

```{r}
my_theme_obs <-  theme(panel.border = element_rect(colour = "black", fill = NA),
                     panel.background = element_blank(),
                     panel.grid = element_blank(),
                     axis.text  = element_text(size = 8), 
                     axis.text.x = element_text(angle = 90, face = "italic"),
                     axis.title = element_text(size = 20),
                     strip.background = element_blank(),
                     strip.text = element_text(size = 10),
                     axis.line.x = element_line(size = 0.5),
                     axis.line.y = element_line(size = 0.5),
                     legend.position = "right",
                     legend.text =element_text(size =6))
```

### Then group, order and plot

```{r}
dat_obs_feat <- dat %>% 
  group_by(SampleID, Type, Depth, Genus.y, depth2) %>% 
  summarise(observed_features = unique(observed_features))

fig7b <- dat_obs_feat %>% 
  drop_na() %>%
  ggplot(aes(Genus.y, observed_features)) +
  geom_point()+
  xlab(NULL)+
  ylab("Observed Features")+
  facet_grid(~depth2, scales = "free_x", space = "free_x", 
             labeller = label_wrap_gen(width = 10))+
  my_theme_obs

fig7b 
#ggsave("figures/observedfeatures-tissue.png", height = 4, width = 5, device= "png")
```
## Merge alpha diversity plots
```{r}
fig7 <- plot_grid(fig7a,fig7b, nrow = 1, rel_heights = 4, rel_widths = 5)

fig7

#ggsave("figures/fig7.png", height = 5, width = 10, device= "png")

```



# Beta Diversity Plots

## Create a PCoA ordination based on unweighted unifrac distances

```{r}
dat_pcoa <- dat %>% 
  group_by(SampleID, PC1 , PC2, Depth, Genus.y, depth2)

dat_pcoa %>%
  filter(Genus.y == "Acropora"| Genus.y == "Galaxea", ) %>%
  drop_na() %>% #drops na values
  #filter(Genus.y != "Acropora")+
  ggplot(aes(x=PC1, y=PC2, color= depth2 , shape = Genus.y))+
  geom_point(alpha=0.5, size = 4) +
  stat_ellipse(level = 0.68)+
  xlab(paste("PC1: ", round(100*uwunifrac$data$ProportionExplained[1]), "%")) +
  ylab(paste("PC2: ", round(100*uwunifrac$data$ProportionExplained[2]), "%"))+
  scale_color_manual(values= c("#0000FF","#771122","#117744","#777711"), name = "Coral Genera")+
  my_theme_alpha

  
  #ggsave("figures/PCoA-tissue.png", height = 4, width = 5, device= "png")
```

# Species Composition plot

## create a theme for the composition plot

```{r}
my_theme_bar <-  theme(panel.border = element_rect(colour = "black", fill = NA),
                     panel.background = element_blank(),
                     panel.grid = element_blank(),
                     axis.text  = element_text(size = 6), 
                     axis.text.x = element_text(angle = 90, face = "italic"),
                     axis.text.x.top = element_text(size=8),
                     axis.title = element_text(size = 12),
                     strip.background = element_blank(),
                     strip.text = element_text(size = 8),
                     axis.line.x = element_line(size = 0.5),
                     axis.line.y = element_line(size = 0.5),
                     legend.position = "right",
                     legend.title = element_text(size = 8),
                     legend.key.size = unit(0.10, "cm"),
                     legend.text =element_text(size =8, face = "italic"))
```

## Create a composition plot by genus

### first lets remove the accession number to then merge the same ITS2 Types with diffrent accession numbers(eg. C1-AB12306.1 and C1-AB12306.2)

```{r}
dat_bar <- dat %>% 
  mutate(ITS2_types= (str_replace_all(pattern = '-.*', replacement = "", Species.x)))

# mutate(species_name2= (sub("\\..*", "", species_name))) %>% 
 # mutate(species_name3= (gsub("(\\d)[a-zA-Z]", "\\1", species_name2)))



```

### Then summurize the top 24 ITS2 sequences

```{r}
total_top_counts <- dat_bar %>% #summarize and arrange the total count of each genus across all samples
  group_by(ITS2_types) %>% 
  summarize(total_top_counts = sum(count)) %>% 
  arrange(desc(total_top_counts))

top_24_species <- total_top_counts %>% #get the 10 most abundant genus by count
  top_n(24,wt=total_top_counts) %>% 
  pull(ITS2_types)

filtered_bar_dat <- dat_bar %>% #filter the data to get the 10 most abundant genera  
  filter(ITS2_types %in% top_24_species)

filtered_bar_dat<- filtered_bar_dat %>% 
  group_by(ITS2_types) %>% 
  mutate(proportion = count / sum(count))

filtered_bar_dat
```

### Reorder the name of the ITS2 types based on abundance and plot the composition graph
##Set colros for ITS2types based on their genus
```{r}
tt <- filtered_bar_dat %>% select(Genus.x, ITS2_types)
nt <- tt %>% count(Genus.x) %>% pivot_wider(names_from = Genus.x, values_from = n)
acols <- colorRampPalette(brewer.pal(9, "Greys"))(nt$Symbiodinium)
ccols <- colorRampPalette(brewer.pal(9, "Greens"))(nt$Symbiodinium)
dcols <- brewer.pal(nt$Durusdinium, "Yellows")




```



```{r}

figure5a <- filtered_bar_dat %>%
  drop_na() %>% 
  filter(Type != "Planula") %>% 
  ggplot(aes(Genus.y, proportion)) +
  facet_grid(~depth2 , scales = "free_x", space = "free_x", labeller = label_wrap_gen(width = 10))+
  geom_bar(aes (fill= ITS2_types), stat = "identity", position = "fill", width = 0.9)+
  scale_fill_discrete(name = NULL)+
  scale_fill_manual(values= c("#0000FF","#00008F","#771122","#771155","#AA4488",
                              "#AA4455","#DD7788","#114477","#4477AA","#77AADD",
                              "#117777","#44AAAA","#117744","#44AA77","#88CCAA",
                              "#77CCCC","#777711","#AAAA44","#AA7744","#DDAA77",
                              "#774411","black","red"), name = "ITS2 Types" )+
  scale_y_continuous(name = "Relative abundance",
                     labels = scales::percent) +
  labs(x = NULL)+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))+
  my_theme_bar

figure5a

#ggsave("figures/species-composition-bar-plot.png", height = 5, width = 10, device= "png")
```
## Complement figure 5a by adding stacked bar plots of individual depth genralist corals (Galaxea and Acropora)
```{r}
figure5b <- filtered_bar_dat %>%
  drop_na() %>% 
  filter(Genus.y == "Acropora", Type != "Planula") %>% 
  ggplot(aes(Genus.y, proportion)) +
  facet_grid(~depth2 , scales = "free_x", space = "free_x", labeller = label_wrap_gen(width = 10))+
  geom_bar(aes (fill= ITS2_types), stat = "identity", position = "fill", width = 0.9)+
  scale_fill_discrete(name = NULL)+
  scale_fill_manual(values= c("#0000FF","#00008F","#771122","#771155","#AA4488",
                              "#AA4455","#DD7788","#114477","#4477AA","#77AADD",
                              "#117777","#44AAAA","#117744","#44AA77","#88CCAA",
                              "#77CCCC","#777711","#AAAA44","#AA7744","#DDAA77",
                              "#774411","black","red"), name = "ITS2 Types" )+
  scale_y_continuous(name = "Relative abundance",
                     labels = scales::percent) +
  labs(x = NULL)+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))+
  my_theme_bar


figure5c <- filtered_bar_dat %>%
  drop_na() %>% 
  filter(Genus.y == "Galaxea") %>% 
  ggplot(aes(Genus.y, proportion)) +
  facet_grid(~depth2 , scales = "free_x", space = "free_x", labeller = label_wrap_gen(width = 10))+
  geom_bar(aes (fill= ITS2_types), stat = "identity", position = "fill", width = 0.9)+
  scale_fill_discrete(name = NULL)+
  scale_fill_manual(values= c("#0000FF","#00008F","#771122","#771155","#AA4488",
                              "#AA4455","#DD7788","#114477","#4477AA","#77AADD",
                              "#117777","#44AAAA","#117744","#44AA77","#88CCAA",
                              "#77CCCC","#777711","#AAAA44","#AA7744","#DDAA77",
                              "#774411","black","red"), name = "ITS2 Types" )+
  scale_y_continuous(name = "Relative abundance",
                     labels = scales::percent) +
  labs(x = NULL)+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))+
  my_theme_bar

figure5 <- plot_grid(figure5a, figure5b,figure5c, rel_widths = 5, rel_heights = 10)
figure5
```




## Bar plot for only A.tenuis samples to compare between Planula and Adult samples

```{r}
Figure6a <- filtered_bar_dat %>% 
  select(SampleID, Type, Species.y, ITS2_types, count) %>% 
  filter(Species.y == "Acropora tenuis") %>% 
  as.data.frame() %>% 
  ggplot(aes(Type, count))+
  geom_bar(aes (fill= ITS2_types), stat = "identity", position = "fill", width = 0.9)+
  scale_fill_discrete(name = NULL)+
  scale_fill_manual(values= c("#0000FF","#00008F","#771122","#771155",
                              "#AA4488","#AA4455","#DD7788","#114477","#4477AA",
                              "#77AADD","#117777","#44AAAA","#117744","#44AA77",
                              "#88CCAA","#77CCCC","#777711","#AAAA44","#DDDD77",
                              "#AA7744","#DDAA77","black", "red"), name = "A.tenuis ITS2 Types" )+
  scale_y_continuous(name = "Relative abundance",
                     labels = scales::percent) +
  labs(x = NULL)+
  my_theme_bar 
  
#ggsave("figures/species-composition-bar-plot-atenuis.png", height = 4, width = 5, device= "png")

```
## Create a pichart showing compositon by coral genera
### lets create a custom theme for this plot too.
```{r}
my_theme_pie <-  theme(panel.border = element_rect(colour = "black", fill = NA),
                     #panel.background = element_blank(),
                     #panel.grid = element_blank(),
                     #axis.text  = element_text(size = 8), 
                     #axis.text.x = element_text(angle = 90, face = "italic"),
                     #axis.text.x.top = element_text(size=8),
                     #axis.title = element_text(size = 16),
                     strip.background = element_blank(),
                     strip.text = element_text(size = 8),
                     axis.line.x = element_line(size = 0.5),
                     axis.line.y = element_line(size = 0.5),
                     legend.position = "right",
                     legend.title = element_text(size = 10),
                     legend.key.size = unit(0.25, "cm"),
                     legend.text =element_text(size =8, face = "italic"))
```


```{r}

filtered_bar_dat %>% 
  filter(! Type== "Planula") %>% 
  ggplot(aes("", count, fill = ITS2_types )) +
  geom_bar(stat = "identity", position = "fill", width = 1)+
  coord_polar("y", start = 0)+
 scale_fill_manual(values= c("#0000FF","#00008F","#771122","#771155",
                              "#AA4488","#AA4455","#DD7788","#114477","#4477AA",
                              "#77AADD","#117777","#44AAAA","#117744","#44AA77",
                              "#88CCAA","#77CCCC","#777711","#AAAA44","#DDDD77",
                              "#AA7744","#DDAA77","black", "red"), name = "ITS2 Types" )+
  scale_y_continuous(name = "Relative abundance",
                     labels = scales::percent) +
  facet_wrap(~Genus.y, labeller=labeller(Genus.y = label_wrap_gen(width = 12)))+
  my_theme_pie +
  theme_void()+
  theme(legend.title = element_text(size = 8),
        legend.text =element_text(size =8, face = "italic"),
        legend.key.size = unit(0.25, "cm"))
  

ggsave("figures/composition-piechart-tissue.png", height = 5, width = 10, device= "png")
```
## Piechart to complement Figure7a

```{r}
figure6b  <- filtered_bar_dat %>% 
  select(SampleID, Type, Species.y, ITS2_types, count) %>% 
  filter(Species.y == "Acropora tenuis") %>% 
  ggplot(aes("", count, fill = ITS2_types )) +
  geom_bar(stat = "identity", position = "fill", width = 1)+
  coord_polar("y", start = 0)+
 scale_fill_manual(values= c("#0000FF","#00008F","#771122","#771155",
                              "#AA4488","#AA4455","#DD7788","#114477","#4477AA",
                              "#77AADD","#117777","#44AAAA","#117744","#44AA77",
                              "#88CCAA","#77CCCC","#777711","#AAAA44","#DDDD77",
                              "#AA7744","#DDAA77","black", "red"), name = "ITS2 Types" )+
  scale_y_continuous(name = "Relative abundance",
                     labels = scales::percent) +
  facet_wrap(~Type, labeller=labeller(Genus.y = label_wrap_gen(width = 12)))+
  my_theme_pie +
  theme_void()+
  theme(legend.position = "none")

figure6b  

## Combine figure b
figure6 <- plot_grid(Figure6a, figure6b, rel_widths = 5, rel_heights = 10)

figure6

ggsave("figures/figure6.png", device= "png")
```






# Summary Analysis

## Counting the most abundant ITSTypes

```{r}
ITS2_sum_by_species <- dat_bar %>% select(FeatureID, SampleID, count, ITS2_types) %>% 
  group_by(ITS2_types) %>% 
  summarize(species_count = sum(count)) %>% 
  mutate(species_percent = species_count/sum(species_count) * 100)

ITS2_sum_by_species
```

### Counting the most abundant taxa after merging ITS2 types with the same names

```{r}
ITS2_merged_sum_by_species <- dat_comp %>% select(FeatureID, SampleID, count, species_name3) %>% 
  group_by(species_name3) %>% 
  summarize(species_count = sum(count)) %>% 
  mutate(species_percent = species_count/sum(species_count) * 100)

ITS2_merged_sum_by_species
```

# Multivariate test analysis using the phyloseq object

## PERMANOVA and betadisper on unweighted unifrac distance

```{r}
uwuni <- phyloseq::distance(physeq, method = "unifrac", weighted = FALSE)  #unweighted unifrac

sampleddf <- data.frame(sample_data(physeq))

adonis2(formula = uwuni ~ Depth * Genus, data = sampleddf)

```
```{r}
betatype <- betadisper(uwuni, sampleddf$Type) #check for sample Type
anova(betatype)
```

```{r}
betagenus <- betadisper(uwuni, sampleddf$Genus)
anova(betagenus)
```
```{r}
TukeyHSD(betagenus)
```


```{r}
betadepth <- betadisper(uwuni, sampleddf$Depth)
anova(betadepth)
```


```{r}
kruskal.test(physeq)
```

#Filter Acropora tenuis samples to compare sample type because our planula are from A. tenuis only.

```{r}
ps_filt_acro <- physeq %>% 
  subset_samples(Species == "Acropora tenuis") %>% 
  subset_samples(sample_sums(physeq) > 0) %>% 
  filter_taxa(function(x) sum(x) > 0, TRUE)
```

```{r}
uwuni_acro <- phyloseq::distance(ps_filt_acro, method = "unifrac", weighted = FALSE)
sampleddf_acro <- data.frame(sample_data(ps_filt_acro))
adonis2(uwuni_acro ~ Type, data = sampleddf_acro)
```

## Try comparing the betdiversity if a sibngle genus found in diffrent depths
### Galaxea
```{r}
ps_filt_gala <- physeq %>% 
  subset_samples(Genus == "Galaxea") %>% 
  subset_samples(sample_sums(physeq) > 0) %>% 
  filter_taxa(function(x) sum(x) > 0, TRUE)

uwuni_gala <- phyloseq::distance(ps_filt_gala, method = "unifrac", weighted = FALSE)
sampleddf_gala <- data.frame(sample_data(ps_filt_gala))
adonis2(uwuni_gala ~ Depth, data = sampleddf_gala)
```

### ACropora

```{r}
ps_filt_acrop <- physeq %>% 
  subset_samples(Genus == "Acropora") %>% 
  subset_samples(sample_sums(physeq) > 0) %>% 
  filter_taxa(function(x) sum(x) > 0, TRUE)

uwuni_acrop <- phyloseq::distance(ps_filt_acrop, method = "unifrac", weighted = FALSE)
sampleddf_acrop <- data.frame(sample_data(ps_filt_acrop))
adonis2(uwuni_acrop ~ Depth, data = sampleddf_acrop)
```


##Making a summary table of ITS2 tyoes by host and depth
```{r}
sum_dat<- dat_bar %>% 
  group_by(ITS2_types) %>% 
  mutate(proportion = count / sum(count))

samples_types <- sum_dat %>%
  filter(!Type == "Planula") %>% 
  # Get clade level relative abundances
  group_by(Species.y, depth2, ITS2_types) %>%
  summarize(Abundance = sum(proportion)) %>%
  group_by(Species.y, depth2) %>%
  ungroup()

significat_types <- samples_types %>% filter(Abundance >= 0.05)

table_final <- significat_types %>% 
  select(Species.y, depth2, ITS2_types) %>% 
  drop_na(ITS2_types) %>% 
  group_by(Species.y, depth2) %>% 
  summarise(taxa = str_c(ITS2_types, collapse = ", "))

#write.csv(table_final, file = "figures/summary_table.csv")
```













































































